---
title: "gg application"
author: "Samantha Walker"
date: "2025-10-30"
output: html_document
---
#stimulate dataset
```{r}
#set seed 
set.seed(137)

stateUniversity <- sample(c("UMN", "UMD", "UMich", "KU", "UGA", "NYU", "Rutgers", "URI", "UPenn", "Ohio", "PennState", "UDel","UConn", "UCLA", "UC Riverside", "UC Santa Barbara", "UC Merced", "NDSU", "USC", "UNC"),100, replace = TRUE)
radius<- abs(rnorm(n = 100, mean = 1, sd=.25))
boba<- sample(0:25, 100, replace=TRUE)
studentSize<- sample(15000:50000,100,replace=TRUE)
schoolClimate<- sample(c("freezing", "snowy","cool", "warm", "hot", "humid"), 100, replace=TRUE)
studentInterest<- sample(c("very low", "low", "moderate", "high", "really high"), 100, replace=TRUE)

bobaDat<-cbind.data.frame(stateUniversity, radius, boba, studentSize, schoolClimate, studentInterest)
```

#### Categorical variable x categorical variable
In the case where we have two categorical predictors, there are a few options we have. We could graph the means: 

load tidyverse
```{r}
library(tidyverse)
```

What are interactions?
- changes in an outcome associated with predictor A that vary as a result of the level of predictor B
- used in regression and ANOVA (which accomplish the same goal w/ different math and assumptions)

Interactions VS Main Effects
- Main Effect: effect of a predictor on the outcome of its own (ignoring the other predictors)
then, can create a table with means 

Categorical Variables
predict # of boba shops based on school climate and student interest
```{r}
stdErr<-function(column){
  sd(column)/sqrt(length(column))
}

summaryBoba<-bobaDat %>% 
  group_by(schoolClimate, studentInterest) %>%
  dplyr::summarise(Mean=mean(boba), stdev=sd(boba), stderr=stdErr(boba))
```
use this output to create a graph 

```{r}
ggplot(summaryBoba, aes(schoolClimate, Mean, fill=studentInterest))+geom_bar(stat="identity", position=position_dodge())+geom_errorbar(aes(ymin=Mean-stderr, ymax=Mean+stderr), position = position_dodge2(), width=0.9)
```

Alternatively, could use facet wrap with boxplots
```{r}
ggplot(bobaDat, aes(schoolClimate, boba, fill=studentInterest))+geom_boxplot()+facet_wrap(~studentInterest)

```

Finally, could use ggeffects package. Need to load the library
```{r}
install.packages("ggeffects")
library(ggeffects)
```

Then, you can use the `ggpredict(model, terms = effects)` function to get predictions. We save these, and then use the `plot()` function to create an interaction plot
```{r}
lm1<-lm(boba~studentInterest*schoolClimate, data=bobaDat)
a<-ggpredict(lm1, terms=c("studentInterest", "schoolClimate"))
plot(a)
```

#### Categorical variable x continuous variable
we've already talked about some ways to show these sorts of interactions

using `facet wrap()`
```{r}
#student Interest and student size
ggplot(bobaDat, aes(studentSize, boba, color=studentInterest))+geom_point()+geom_smooth(method="lm")+facet_wrap(~studentInterest)
```

using color 
```{r}

```

using ggeffects package
```{r}

```

#### continuous x continuous interaction
using ggeffects package
```{r}
lm2<-lm(boba~studentSize*studentInterest, data=bobaDat)
a2<-ggpredict(lm2)
plot(a2)

lm2<-lm(boba~studentSize*studentInterest, data=bobaDat)
a2<-ggpredict(lm2)
plot(a2)

ggplot(bobaDat, studentSize, radius)

```


### Learning Outcome 2: Identify Code to create a shiny app 
## Before we get started
Shiny apps are advanced. Even if we're focusing on "basics" the coding behind it is more advanced than what we covered last week. It's **ok** if you walk away from the lesson not knowing what I'm talking about. 

The plan is to **show** you what you can build to as you program in R. There are tutorials that are ~18 hours to complete. We can't do that in one class. I've included links that I use when making shiny apps that you may want to try. 

#### Identify Code to create a shiny app 
shiny is a package in R that allows us to create interactive graphs. You can create dashboards and place interactive graphs online. 

Today should be treated as an introduction. You can learn more here: https://shiny.rstudio.com/tutorial/ 

People make shiny apps in different ways, I'm going to use naming conventions and descriptions used by R Studio developers because it makes it easier to follow along. 

But first, we need to load the shiny package. 
```{r}
library(shiny)
```

Before we start coding, we need to talk about the different aspects of a shiny app: 
**User Interface:** We're programming the part that someone would see in the plot
- If you place it on the internet, this would be the portion that others would see 
**Server:** This is the background code to create what the user sees. It's what the computer runs in the background. (ie R code)  

The **User Interface** creates the webpage, and the **server** tells R how to put the pieces together.

Then, **shinyApp** combines the information in the User interface and the Server. We'll start with an example below to get an idea of the structure. 

Just like we started with ggplot, we'll start with a shell. 
```{r}
ui <- fluidPage()
server <- function(input, output) {}
shinyApp(ui = ui, server = server)
```
I am using ui to stand for the user interface. This is going to be what someone would see. Notice that the server is a function that we are designing, and we're giving it inputs and outputs. In this case, it is an empty function. 

Whatever we put in the fluidpage() function above is going to be present in our shiny app. 
```{r}
ui <- fluidPage("I LOVE PSY4802!")
server <- function(input, output) {}
shinyApp(ui = ui, server = server)
```
You can see that whatever I write in fluid page, shows up in the app. When we want to add more to what other people will see, we need to specify different inputs and outputs.   

In this section, we will talk about different inputs and outputs you can use. 

**Creating a scale**
Making graphs is complex. We're going to start with making a scale input because it helps us understand how the input functions work.  

All inputs follow similar structures, so you will be able to apply this general format to other options.

The function to make a scale is `sliderInput()` and takes the following inputs: 
`sliderInput(inputId, label, min, max, value)`

Just looking at the input code (not in the shiny app), it will look something like this: 
```{r}
sliderInput(inputId="number", label="Pick a number", value=13, min=1, max=100)
```

When we put it in the ui/server code we used to make the app: 
```{r}
ui <- fluidPage(
  sliderInput(inputId = "number", 
    label = "Pick a number", 
    value = 13, min = 0, max = 100),
  plotOutput("hist")
)

server <- function(input, output) {}

shinyApp(ui = ui, server = server)
```
This code lets a person using the Shiny app pick a number between 1 and 100. We can then use this to create an output that will change based on people's responses. 

**Creating a graph**
In this example, we'll create a histogram that plots a random normal sample of size n (whatever the person picks).

Before we look at the shiny app code, I need to walk you through the code we'll use as an output. 

first, we'll use rnorm. We can use this function to draw a random sample from a normal distribution (mean= 0, sd=1). 
```{r}
set.seed(25)
rnorm(32)
```
We'll put this into a plot (which will let me show you a way to graph vectors not in a dataframe)
```{r}
library(ggplot2)
ggplot(,aes(rnorm(32)))+ geom_histogram()
```

```{r}
ui <- fluidPage(
  sliderInput(inputId = "number", 
    label = "Pick a number", 
    value = 13, min = 0, max = 100),
  plotOutput("hist")
)

server <- function(input, output) {
  output$hist <- renderPlot({
    ggplot(, aes(rnorm(input$number)))+ geom_histogram(fill="darksalmon")+theme_classic()
  })
}

shinyApp(ui = ui, server = server)
```
#### Practice (Challenge)
1. Create a shiny app that allows people to select a number between 1 and 5000. 
```{r}
ui <- fluidPage(
  sliderInput(inputId = "number", 
    label = "Pick a number", 
    value = 2500, min = 1, max = 5000))
server <- function(input, output) {}

shinyApp(ui = ui, server = server)

```

2. Add an output that creates a graph based on these numbers.
```{r}
ui <- fluidPage(
  sliderInput(inputId = "number", 
    label = "Pick a number", 
    value = 2500, min = 1, max = 5000), plotOutput("hist")
)

server <- function(input, output) {
  output$hist <- renderPlot({
    ggplot(, aes(rnorm(input$number)))+ geom_histogram(fill="midnightblue")+theme_classic()
  })
}

shinyApp(ui = ui, server = server)

```

```{r}
ui <- fluidPage(
  sliderInput(inputId = "number", 
    label = "Pick a number", 
    value = 2500, min = 1, max = 5000), plotOutput("boxp")
)

server <- function(input, output) {
  output$boxp <- renderPlot({
    ggplot(, aes(rnorm(input$number)))+ geom_boxplot(fill="midnightblue")+theme_classic()
  })
}

shinyApp(ui = ui, server = server)

```
**Challenge:** Create a graph based on two numbers chosen by the user
```{r}
ui <- fluidPage(
  sliderInput(inputId = "number1", 
    label = "Pick a number", 
    value = 2500, min = 1, max = 5000),sliderInput(inputId = "number2", 
    label = "Pick a number", 
    value = 2500, min = 1, max = 5000), plotOutput("scatterplot")
)

server <- function(input, output) {
  output$scatterplot <- renderPlot({
    ggplot(, aes(rnorm(input$number1), rnorm(input$number2)))+ geom_point(color="purple")+theme_classic()
  })
}

shinyApp(ui = ui, server = server)


```

### Using real data in a shiny app

For this graph, we're going to use built in R data and allow users to choose what they'd like to graph. To do this, we'll use a dropdown list as the input. (for other inputs: https://shiny.rstudio.com/images/shiny-cheatsheet.pdf)

Just like we did before, I'll show you what code we're adding to the ui side before we add it. 

We're going to use the diamonds dataset for this practice
```{r}
data(diamonds)
```

**User Interface Set up**
The first part we'll set up is the ui. We'll be making a drop down list. 

To make a dropdown list, we'll use the selectInputs function. It takes the following arguments:
`selectInput(inputId, label (what other people see), choices(that people can choose from))`
```{r}
ui<- fluidPage(
  titlePanel("diamonds data"),
  sidebarLayout(
    sidebarPanel(
  selectInput(inputId= "colorChoice", label="Select Diamond Color", choices= c("D"="D", "E"="E", "F"= "F", "G"="G", "H"="H", "I"="I", "J"="J"))),
  mainPanel(
plotOutput("colorgraph")
)
)
)
server <- function(input, output) {
  output$colorgraph <- renderPlot({
    ggplot(diamonds[diamonds$color==input$colorChoice,], aes(carat))+ geom_histogram(fill= "lightseagreen",bins= 50)+theme_classic()
  })
}


shinyApp(ui = ui, server = server)
```

```{r}
ui<- fluidPage(
  titlePanel("diamonds data"),
  sidebarLayout(
    sidebarPanel(
  selectInput(inputId= "colorChoice", label="Select Diamond Color", choices= c("D"="D", "E"="E", "F"= "F", "G"="G", "H"="H", "I"="I", "J"="J"))),
  mainPanel(
plotOutput("colorgraph")
)
)
)
server <- function(input, output) {
  output$colorgraph <- renderPlot({
    ggplot(diamonds[diamonds$color==input$colorChoice,], aes(carat, price))+ geom_point(color= "lightseagreen")+theme_classic()+ labs(title= "carat size of diamonds")
  })
}

shinyApp(ui = ui, server = server)
```
