---
title: "ggplot pt. 2"
author: "Samantha Walker"
date: "2025-10-28"
output: html_document
---

# Learning Outcomes
By the end of this class, students should be able to: 
- Explain how to alter information in the legend 
- Describe how to use facet wrap
- Explain how to export graphs


### Learning Outcome 1: Explain how to alter information in the legend
It can also be helpful to understand how to modify your legend: 
`+ theme(legend.position="none")` will remove the legend. You can also choose to move the legend to the top or bottom of the graph

Let's try this on a graph from the diamonds dataset: 
```{r}
library(ggplot2)
data(diamonds)
ggplot(diamonds, aes(color, price, fill=cut))+geom_boxplot()
```

We can also change the location: 
```{r}
ggplot(diamonds, aes(color, price, fill=cut))+geom_boxplot()+theme(legend.position="bottom")
#top, left, right, bottom, or none are the options

```

To Change the legend title, you can use a couple of different codes, depending on the type of data: 
`scale_color_manual(name="Title")` works for scatter plots filled in with a factor
`scale_fill_manual(name="Title)` should work for a bar graph filled in by a factor
`scale_color_continuous()` and `scale_fill_continuous()` should work for a numeric fill
```{r}
ggplot(diamonds, aes(color, price, fill=cut))+geom_boxplot()+theme(legend.position="bottom")+scale_fill_manual(name="Cut of Diamonds", values=c("purple", "blue", "green", "orange", "red", "yellow", "pink")
```

### Practice:

1) make a graph of bill length and bill depth. Include island as a third variable
```{r}
library(palmerpenguins)
data("penguins")
ggplot(penguins, aes(bill_len, bill_dep, fill=island))+geom_point()
```

2) Move the legend to the top of the graph
```{r}
ggplot(penguins, aes(bill_length_mm, bill_depth_mm, fill=island))+geom_point()+theme(legend.position = "top")
```

3) Remove the legend from the graph 
```{r}
ggplot(penguins, aes(bill_len, bill_dep, fill=island))+geom_point()+theme(legend.position = "none")

```

4) Change the title of the legend
```{r}
ggplot(penguins, aes(bill_length_mm, bill_depth_mm, color=island))+geom_point()+theme(legend.position = "top")+scale_fill_manual(name="Island", values=c("pink", "orange", "yellow"))

```


5) make a graph of bill length and bill depth. Include flipper length as a third variable
```{r}
ggplot(penguins, aes(bill_length_mm, bill_depth_mm, cut=flipper_length_mm))+geom_point()

### Learning Outcome 2: Describe how to use facet wrap
Facet Wrap allows us to incorporate a third variable in another way. Specifically, it panels graphs so that we can make an individual graph for each level of a factor.

we need to make a graph and save it 
```{r}
data(diamonds)
ggplot(diamonds, aes(color, price, color=cut))+geom_point()

ggplot(diamonds, aes(color, price, color=cut))+geom_boxplot()+facet_wrap(~cut)

ggplot(diamonds, aes(carat, price, color=cut))+geom_point()+geom_smooth(method="lm")+facet_wrap(~cut)

ggplot(diamonds, aes(carat, price, color=cut))+geom_point()+geom_smooth(method="lm")+facet_wrap(~color)
#creates a graph for each cut of diamond
```

### change labels on the axes
change levels beforehand
```{r}
levels(penguins$species)<-c("Adelie Penguin", "Chinstrap Penguin", "Gentoo Penguin")

ggplot(penguins, aes(bill_length_mm, bill_depth_mm))+geom_point()+facet_wrap(~species)
```

make a vector called labels. use `labeller=as_labeller())` in `facet_wrap()`
```{r}
speciesTitle<-c("Adelie Penguin"="Adelia", "Chinstrap Penguin" ="Chinstrap", "Gentoo Penguin"="Gentoo")
ggplot(penguins, aes(bill_length_mm, bill_depth_mm))+geom_point()+facet_wrap(~species, labeller = as_labeller(speciesTitle))
```

### Practice: 
1) create a facet wrap graph to generate a scatterplot of mpg and hp based on the number of cylinders a car has
```{r}
data("mtcars")
cyl_title<-c("4"="4 cyl", "6"="6 cyl", "8"="8 cyl")
ggplot(mtcars, aes(hp, mpg, color=cyl))+geom_point()+facet_wrap(~cyl, labeller = as_labeller(cyl_title))
```

### Learning Outcome 3: Explain how to export graphs
to export graphs, I use the command `ggsave()`
Specifically, you can use the following arguments: 
`ggsave("name of graph", width, height, units, dpi)`
```{r}
SexLabels<-c("female"="FEMALE", "male"="MALE", "NA"="Unknown")
ggplot(penguins, aes(bill_length_mm, bill_depth_mm, color=sex))+geom_point()+facet_wrap(~sex, labeller = as_labeller(SexLabels))+labs(title="Bill Length and Depth by Gender", x="Bill Length", y="Bill Depth")+theme(legend.position = "bottom")

ggsave("FinalPrettyPenguinGraph.tiff", width=10, height=8, units="in", dpi=300)
```
I don't usually specify a path in ggsave, but make sure my r code is in the directory I'd like it to be saved in. you can check your working directory by using this code
```{r}
getwd()
```
you can change your working directory using this code: 
```{r}
setwd()
```